The truncated KdV framework for modeling anomalous waves induced by abrupt depth changes

### 2019-07-12 WavesData-2019/LatexNotes/notes.tex
NM

% Basic editing
\newcommand{\tocite}{{\color{blue}(to cite)}}
\newcommand{\vsp}[1]{\vspace{#1 pc} \noindent}
\newcommand{\np}{\newpage \noindent}
% Derivatives
\newcommand{\pd}[2]    { \frac{\partial #1} {\partial #2} }
\newcommand{\ppd}[2]  { \frac{\partial^2 #1}{{\partial #2}^2} }
\newcommand{\pdi}[2] { {\partial_#2} #1 }
\newcommand{\td}[2] { \frac{d #1} { d #2 } }
\newcommand{\grad}{\nabla}
\newcommand {\Lap} {\grad^2}
% Vectors and operators
\newcommand{\bvec}[1]{\ensuremath{\boldsymbol{#1}}}
\newcommand{\abs}[1]{\left| #1 \right|}
\newcommand{\norm}[1]{\left\| #1 \right\|}
\newcommand{\mean}[1]{\left< #1 \right>}
\newcommand{\eps}{\varepsilon}

% Basic physical parameters and scales
\newcommand{\depth}{h}
\newcommand{\dup}{\depth^{-}}
\newcommand{\ddn}{\depth^{+}}
\newcommand{\freqp}{f_p}
\newcommand{\lam}{\lambda}
\newcommand{\lamup}{\lam^{-}}
\newcommand{\lamfac}{N}
% More parameters
\newcommand{\etastd}{\eta_{\text std}}
\newcommand{\amp}{a}
% Statistical quantities
\newcommand{\skw}{\text{skew}}
\newcommand{\var}{\text{var}}
% Dimensionless parameters
\newcommand{\epsup}{\eps_0}
\newcommand{\delup}{\delta_0}
\newcommand{\drat}{D}
\newcommand{\ampp}{\mathcal{A}}
% Hamiltonian and Gibbs stuff
\newcommand{\Hp}{H^{+}}
\newcommand{\Hm}{H^{-}}
\newcommand{\Gibbs}{\mathcal{G}}
\newcommand{\Gp}{\Gibbs^{+}}
\newcommand{\Gm}{\Gibbs^{-}}
\newcommand{\meanup}[1]{\mean{#1}_{-}}
\newcommand{\meandn}[1]{\mean{#1}_{+}}

@@

The dimensionless KdV equation takes the form
\begin{equation}
u_t + \frac{3 \pi}{\lamfac} \epsup \drat^{-1} \, u u_x + \frac{4 \pi^3}{3 \lamfac^3} \delup^2 \drat \, u_{xxx} = 0
\qquad \text{for } x \in [-\pi,\pi]
\end{equation}
where dimensionless parameters are as follows
\begin{align}
&\epsup = \amp/\dup \, , \\
&\delup = \dup/\lamup = \sqrt{\dup \freqp^2/g} \, \\
&\drat(x) = {\depth(x)}/{\dup} \, .
\end{align}
$\drat = \drat(x)$ is the local dimensionless depth.

@@

@@

Note, that as a consequence of the definition of $\amp$ in \eqref{ampdefn}, we have that
\begin{equation}
E[u] = \frac{1}{2} \frac{1}{2 \pi} \int_{-\pi}^{\pi} u^2 \dx = \frac{1}{2}
\end{equation}

@@

% Table
%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
\begin{table}[h]%[htbp]
\begin{center}
\caption{Table of parameters}
\label{paramtable}
\begin{tabular}{l l l l}
\hline \multicolumn{4} { c }{Parameters that are constant throughout the domain} \\
\hline Description & Notation and definition & Units & Value in experiments \\
\hline
Peak forcing frequency		& $f_p$						& frequency	& 2 Hz \\
Characteristic wave amplitude	& $a = \mean{\eta^2}^{1/2} = \etastd$		& length		& 0.03--0.3 cm \\
Upstream wavelength			& $\lambda_0 = \sqrt{g h_0}/f_p$	& length		& 56 cm \\
Amplitude-to-depth ratio		& $\eps = a / h_0$				& dimensionless	& 0.0024--0.024 \\
Depth-to-wavelength ratio		& $\delta = h_0 / \lambda_0$		& dimensionless		& 0.22 \\
Amplitude parameter			& $\ampp = \eps \delta^{-2} = a \lambda_0^2 h_0^{-3} = ag h_0^{-2} f_p^{-2}$	
& dimensionless		& 0.05--0.5\\
Catch-all parameter		& $\mu = 4 \pi^2 / (9 \lamfac^2)$			& dimensionless		& 0.27 if $\lamfac=4$ \\
\hline \multicolumn{4} { c }{Parameters that change value across ADC} \\
\hline Description & Notation and definition & Units \\
\hline
Local depth			& $d$					& length \\
Local wavespeed		& $c = \sqrt{gd}$			& speed \\
Local peak wavelength	& $\lambda_p = \sqrt{gd}/f_p$	& length \\
Characteristic wavelength	& $\lambda_c = \lamfac \lambda_p$	& length \\
Depth ratio			& $D = d/h_0$				& dimensionless
\end{tabular}
\end{center}
\end{table}
 %^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%

>>>>

% Basic editing
\newcommand{\tocite}{{\color{blue}(to cite)}}
\newcommand{\vsp}[1]{\vspace{#1 pc} \noindent}
\newcommand{\np}{\newpage \noindent}
% Derivatives
\newcommand{\pd}[2]    { \frac{\partial #1} {\partial #2} }
\newcommand{\ppd}[2]  { \frac{\partial^2 #1}{{\partial #2}^2} }
\newcommand{\pdi}[2] { {\partial_#2} #1 }
\newcommand{\td}[2] { \frac{d #1} { d #2 } }
\newcommand{\grad}{\nabla}
\newcommand {\Lap} {\grad^2}
% Vectors and operators
\newcommand{\bvec}[1]{\ensuremath{\boldsymbol{#1}}}
\newcommand{\abs}[1]{\left| #1 \right|}
\newcommand{\norm}[1]{\left\| #1 \right\|}
\newcommand{\mean}[1]{\left< #1 \right>}
\newcommand{\eps}{\varepsilon}
\newcommand{\defeq}{\mathrel{\mathop:}=}

% Basic physical parameters and scales
\newcommand{\depth}{h}
\newcommand{\dup}{\depth_{-}}
\newcommand{\ddn}{\depth_{+}}
\newcommand{\freqp}{f_p}
\newcommand{\lam}{\lambda}
\newcommand{\lamup}{\lam_{-}}
\newcommand{\lamfac}{N}
% More parameters
\newcommand{\etastd}{\eta_{\text std}}
\newcommand{\amp}{a}
% Statistical quantities
\newcommand{\skw}{\text{skew}}
\newcommand{\var}{\text{var}}
% Dimensionless parameters
\newcommand{\epsup}{\eps_0}
\newcommand{\delup}{\delta_0}
\newcommand{\drat}{D}
\newcommand{\ampp}{\mathcal{A}}
% Hamiltonian and Gibbs stuff
\newcommand{\Hp}{H^{+}}
\newcommand{\Hm}{H^{-}}
\newcommand{\Gibbs}{\mathcal{G}}
\newcommand{\Gp}{\Gibbs^{+}}
\newcommand{\Gm}{\Gibbs^{-}}
\newcommand{\meanup}[1]{\mean{#1}_{-}}
\newcommand{\meandn}[1]{\mean{#1}_{+}}

@@

Dropping the tildes, the dimensionless KdV equation takes the form
\begin{equation}
u_t + \frac{3 \pi}{\lamfac} \epsup \drat^{-1} \, u u_x + \frac{4 \pi^3}{3 \lamfac^3} \delup^2 \drat \, u_{xxx} = 0
\qquad \text{for } x \in [-\pi,\pi]
\end{equation}
where dimensionless variables are as follows
\begin{align}
&\epsup = \amp/\dup \, , \\
&\delup = \dup/\lamup = \sqrt{\dup \freqp^2/g} \, \\
&\drat(x) = {\depth(x)}/{\dup} \, .
\end{align}
Here, $\drat = \drat(x)$ is the local dimensionless depth.

@@

% Table
%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
\begin{table}[h]%[htbp]
\begin{center}
\caption{Table of parameters}
\label{paramtable}
\begin{tabular}{l l l l}
\hline \multicolumn{4} { c }{Parameters that are constant throughout the domain} \\
\hline Description & Notation and definition & Units & Value in experiments \\
\hline
Peak forcing frequency		& $f_p$						& frequency	& 2 Hz \\
Characteristic wave amplitude	& $a = \mean{\eta^2}^{1/2} = \etastd$		& length		& 0.03--0.3 cm \\
Upstream wavelength			& $\lamup = \sqrt{g \dup}/f_p$	& length		& 56 cm \\
Amplitude-to-depth ratio		& $\epsup = a / \dup$			& dimensionless	& 0.0024--0.024 \\
Depth-to-wavelength ratio		& $\delup = \dup / \lamup$		& dimensionless		& 0.22 \\
%Amplitude parameter			& $\ampp = \eps \delta^{-2} = a \lambda_0^2 h_0^{-3} = ag h_0^{-2} f_p^{-2}$	& dimensionless		& 0.05--0.5\\
%Catch-all parameter		& $\mu = 4 \pi^2 / (9 \lamfac^2)$			& dimensionless		& 0.27 if $\lamfac=4$ \\
\hline \multicolumn{4} { c }{Parameters that change value across ADC} \\
\hline Description & Notation and definition & Units \\
\hline
Local depth			& $\depth$				& length \\
Local wavespeed		& $c = \sqrt{g \depth}$		& speed \\
Local peak wavelength	& $\lam = \sqrt{gd}/f_p$	& length \\
%Characteristic wavelength	& $\lambda_c = \lamfac \lambda_p$	& length \\
Depth ratio			& $\drat = \depth/\dup$		& dimensionless
\end{tabular}
\end{center}
\end{table}
 %^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%

\subsubsection{Alternative KdV formulation following Johnson}
The above makes the implicit assumption of continuous surface displacement crossing the ADC. This is a very physically motivated assumption, but there is a small amount of debate surrounding the assumption. An alternative is to appeal to Green's law and assume the quantity $\depth^{1/4} \eta$ to be continuous across the ADC, which necessarily implies a small discontinuity in the surface (which of course cannot be true literally). The dimensional KdV equation would then be expressed as
\begin{equation}
2 (\depth^{1/4} \eta)_t + \frac{3 c}{\depth^{5/4}} (\depth^{1/4} \eta) (\depth^{1/4} \eta)_{\xi} + \frac{c \depth^2}{3} (\depth^{1/4} \eta)_{\xi \xi \xi} = 0
\end{equation}
This is the approach taken in Johnson's book. This assumption results in a small modification of the coefficients in the dimensionless KdV. In particular, it gives
\begin{equation}
u_t + \frac{3 \pi}{\lamfac} \epsup \drat^{-5/4} \, u u_x + \frac{4 \pi^3}{3 \lamfac^3} \delup^2 \drat \, u_{xxx} = 0
\qquad \text{for } x \in [-\pi,\pi]
\end{equation}
where the state variable is
\begin{equation}
u = \frac{\depth^{1/4} \, \eta}{ \dup^{1/4} \, \amp}
\end{equation}
Notice the only changes are the definition of the state variable and the small change in the power of $\drat$ in the second term (changes from $\drat^{-1}$ to $\drat^{-5/4}$).

@@

The domain of consideration is $x \in [-\pi,\pi]$. We first define the spatial average of any quantity as
\begin{equation}
\mean{q} = \frac{1}{2\pi} \int_{-\pi}^{\pi} q \dx \, .
\end{equation}
Note that from the definition of the characteristic amplitude in \eqref{ampdefn}, we have $\mean{u^2} = 1$. Consequently, the energy of $u$ is fixed as
\begin{equation}
E[u] \defeq \frac{1}{2} \int_{-\pi}^{\pi} u^2 \dx = \pi
\end{equation}

@@
